### 2025-03-17 paper/preprint/paper.tex

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-default}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-warning-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-warning-color!10!white, title={Package Version (Reproducibility)}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

Tuning was run using \texttt{v1.1.3} of \texttt{TaijaData}. The
follow-up version \texttt{v1.1.4} introduced an option to split
real-world tabular datasets into train and test set, ensuring that
pre-processing steps like standardization is fit on the training set
only. If you are rerunning the tuning experiments with a version of
\texttt{TaijaData} that is higher than \texttt{v1.1.3}, than for the
default parameters specified in the configuration files, you may end up
with slightly different results, although we would not expect any
changes in terms of qualitative findings. For exact reproducibility,
please use \texttt{v1.1.3}.

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

@@

\begin{tcolorbox}[enhanced jigsaw, colframe=quarto-callout-note-color-frame, opacityback=0, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colback=white, arc=.35mm, opacitybacktitle=0.6, bottomtitle=1mm, breakable, titlerule=0mm, toptitle=1mm, rightrule=.15mm, bottomrule=.15mm, toprule=.15mm, left=2mm, leftrule=.75mm]

>>>>

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-default}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-eval-default}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-gen-params-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-pen-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-train-final-run-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-train}: Training Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

\begin{tcolorbox}[enhanced jigsaw, title={Note \ref*{nte-tune_lr-eval}: Evaluation Phase}, colframe=quarto-callout-note-color-frame, bottomrule=.15mm, colback=white, opacitybacktitle=0.6, bottomtitle=1mm, colbacktitle=quarto-callout-note-color!10!white, coltitle=black, breakable, toptitle=1mm, arc=.35mm, left=2mm, titlerule=0mm, rightrule=.15mm, toprule=.15mm, leftrule=.75mm, opacityback=0]

### 2025-03-17 paper/sections/abstract.qmd

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be (1) plausible with respect to the underlying data and (2) actionable with respect to the user-defined mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be (1) plausible with respect to the underlying data and (2) actionable with respect to the user-defined mutability constraints. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while promoting robustness and preserving high predictive performance.

### 2025-03-17 paper/sections/intro.qmd

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models [@mcgregor2021preventing; @oneil2016weapons]. Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

@@

 The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability without sacrificing predictive performance. We discuss the challenges in @sec-discussion and conclude in @sec-conclusion that CT is a promising approach towards making opaque models more trustworthy.

>>>>

The first challenge has attracted an abundance of research on **explainable AI** (XAI), a paradigm that focuses on the development of tools to derive (post-hoc) explanations from complex model representations. Such explanations should mitigate a scenario in which practitioners deploy opaque models and blindly rely on their predictions. On countless occasions, this has happened in practice and caused real harms to people who were adversely and unfairly affected by automated decision-making (ADM) systems involving opaque models (see, e.g., [@oneil2016weapons]). Effective XAI tools can aid us in monitoring models and providing recourse to individuals to turn negative outcomes (e.g., "loan application rejected") into positive ones (e.g., "application accepted"). In line with this, our work builds upon **counterfactual explanations** (CE) proposed by @wachter2017counterfactual as an effective approach to achieve this goal. CEs prescribe minimal changes for factual inputs that, if implemented, would prompt some fitted model to produce a desired output.

@@

 The remainder of this manuscript is structured as follows. @sec-lit presents related work, focusing on the link between AEs and CEs. Then follow our two principal contributions. In @sec-method, we introduce our methodological framework and show theoretically that it can be employed to enforce global actionability constraints. In @sec-experiments, through extensive experiments, we demonstrate that CT substantially improves explainability and positively contributes to the robustness of trained models without sacrificing predictive performance. Finally, in @sec-discussion, we discuss the challenges and conclude that CT is a promising approach towards making opaque models more trustworthy.

### 2025-03-17 paper/sections/other/training.qmd

## Training Details {#sec-app-training}

In this section, we describe the training procedure in detail. While the details laid out here are not crucial for understanding our proposed approach, they are of importance to anyone looking to implement counterfactual training.

### Default Hyperparameters

@@

@nte-eval-default presents the default hyperparameters used during evaluation.

>>>>

## Training Hyperparameters {#sec-app-training}

@@

For all of our evaluations, we proceed as follows: for each experiment setting we generate multiple counterfactuals ("No. Counterfactuals"), randomly choosing the factual and target class each time (@nte-eval-default). We do this across multiple rounds ("No. Runs") with different random seeds to account for stochasticity (@nte-eval-default). This is in line with standard practice in the related literature on CE. @nte-eval-default presents the default hyperparameters used during evaluation. For our final results presented in the main paper, we rely on held out test sets to sample factuals (and outputs for our performance metrics). For tuning purposes we rely on training or validation sets.

### Robust Accuracy

To evaluate robust accuracy (Acc.$^*$), we use the Fast Gradient Sign Method (FGSM) to perturb test samples [@goodfellow2014explaining]. For the main results, we have set the perturbation size to $\epsilon=0.03$. We have also tested other perturbation sizes, as well as randomly perturbed data. Although not reported here, we have consistently found strong outperformance of CT compared to the weak baseline.
